---
title: "Getting Started with claudeR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with claudeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

claudeR provides a compact R interface to Anthropic's Messages API. It accepts
simple string prompts or complete message lists, supports adaptive thinking and
streaming, exposes structured stop metadata, and can query the live Models API.
The package also includes a separate interface to the Claude Code CLI.

## Prerequisites

Create an Anthropic API key at <https://console.anthropic.com/> and make it
available as `ANTHROPIC_API_KEY`:

```{r}
Sys.setenv(ANTHROPIC_API_KEY = "your-api-key-here")
```

For persistent use, add the following line to `.Renviron` instead:

```
ANTHROPIC_API_KEY=your-api-key-here
```

Load the package:

```{r setup}
library(claudeR)
```

## Basic Usage

A single string is automatically wrapped as a user message. The default
`model = "sonnet"` shortcut selects Claude Sonnet 5.

```{r}
response <- claudeR(
  "What is the capital of France?",
  max_tokens = 100
)
print(response)
```

Add top-level system instructions with `system_prompt`:

```{r}
response <- claudeR(
  "Translate: Hello, how are you?",
  system_prompt = "You are a Spanish translator. Return only the translation.",
  max_tokens = 100
)
print(response)
```

For multi-turn or multimodal requests, provide complete Messages API objects:

```{r}
response <- claudeR(
  prompt = list(
    list(role = "user", content = "My name is Alice."),
    list(role = "assistant", content = "Hello Alice!"),
    list(role = "user", content = "What is my name?")
  ),
  max_tokens = 100
)
print(response)
```

## Current Models

claudeR defines four case-insensitive convenience shortcuts:

| Model | Shortcut | API model ID | Typical use |
|-------|----------|--------------|-------------|
| Claude Fable 5 | `"fable"` | `claude-fable-5` | Highest-capability, long-running work |
| Claude Opus 4.8 | `"opus"` | `claude-opus-4-8` | Complex agentic coding and knowledge work |
| Claude Sonnet 5 | `"sonnet"` | `claude-sonnet-5` | Balanced speed and capability; default |
| Claude Haiku 4.5 | `"haiku"` | `claude-haiku-4-5-20251001` | Fast, cost-efficient tasks |

Full model IDs are also accepted, including IDs released after this package
version. Starting with Claude 4.6, a dateless ID such as
`claude-opus-4-8` is a canonical, pinned version rather than an evergreen alias.
The Haiku shortcut resolves to its dated, pinned ID; Anthropic's shorter
`claude-haiku-4-5` API alias is also accepted when supplied explicitly.

Use `claude_models()` to inspect live availability and capabilities rather than
depending only on a static table:

```{r}
models <- claude_models(limit = 25)
vapply(models$data, `[[`, character(1), "id")

# A shortcut is resolved when retrieving a single model
opus <- claude_models("opus")
opus$id
opus$max_input_tokens
opus$capabilities
```

## Adaptive Thinking and Effort

The current Fable, Opus, and Sonnet models use adaptive thinking. Use `effort`
to tune reasoning depth; claudeR sends it as `output_config$effort`.

```{r}
result <- claudeR(
  "Find edge cases in an R function that computes grouped weighted means.",
  model = "opus",
  thinking = list(type = "adaptive", display = "summarized"),
  effort = "high",
  return_thinking = TRUE,
  max_tokens = 4000
)

cat("Thinking summary:\n", result$thinking, "\n\n")
cat("Answer:\n", result$response)
```

Do not use a fixed `budget_tokens` configuration with current adaptive models.
Also leave `temperature`, `top_k`, and `top_p` as `NULL`; modern adaptive models
reject non-default sampling values.

## Fable 5 Refusals

Fable 5 always uses adaptive thinking, so omit `thinking`; explicitly disabling
thinking is invalid. A safety classifier can decline a Fable request with an
HTTP 200 response and `stop_reason = "refusal"`. Request the structured response
to distinguish a refusal from an empty answer:

```{r}
result <- claudeR(
  "Plan a long-running research workflow.",
  model = "fable",
  effort = "high",
  return_response = TRUE,
  max_tokens = 4000
)

if (identical(result$stop_reason, "refusal")) {
  detail <- result$stop_details$explanation
  if (is.null(detail)) detail <- "Claude declined the request."
  message(detail)
} else {
  cat(result$response)
}
```

With `return_response = TRUE`, the result also includes content blocks, model,
usage, stop sequence, and stop details. Without it, claudeR warns and returns
`NULL` for a refusal.

## Beta Headers and Fallbacks

Supply Anthropic beta header names with `betas`. Additional named Messages API
fields pass through `...`. This example uses the current server-side fallback
beta to retry a Fable refusal with Opus 4.8:

```{r}
result <- claudeR(
  "Analyze this difficult task.",
  model = "fable",
  return_response = TRUE,
  betas = "server-side-fallback-2026-06-01",
  fallbacks = list(list(model = "claude-opus-4-8"))
)
```

Beta names and availability can change. Check Anthropic's current documentation
before production use. Values nested in passthrough fields, such as a fallback
model, must use full API IDs rather than claudeR shortcuts.

## Streaming and Return Values

`stream = TRUE` streams response text as it arrives. Use
`return_thinking = TRUE` to return extracted thinking and response strings, or
`return_response = TRUE` for content blocks and complete stop metadata. The
historical `stream_thinking` argument remains as a compatibility alias.

```{r}
result <- claudeR(
  "Explain how lexical scoping works in R.",
  stream = TRUE,
  return_response = TRUE
)
result$usage
```

## Main Parameters

| Parameter | Description | Default |
|-----------|-------------|---------|
| `prompt` | One string or a non-empty list of message objects | required |
| `model` | Full API model ID or claudeR shortcut | `"sonnet"` |
| `max_tokens` | Maximum output tokens | `1024` |
| `system_prompt` | Top-level system instructions | `NULL` |
| `thinking` | Adaptive-thinking configuration | `NULL` |
| `effort` | Reasoning effort under `output_config` | `NULL` |
| `stream` | Stream the response | `FALSE` |
| `return_thinking` | Return thinking and response strings | `FALSE` |
| `return_response` | Return structured response and stop metadata | `FALSE` |
| `betas` | Anthropic beta header names | `NULL` |
| `...` | Additional named Messages API fields | none |

API validation failures and non-success HTTP responses raise R errors. Fable
classifier refusals are successful API responses and follow the special handling
described above.

## Claude Code Model Selection

`claude_code()` invokes the separately installed Claude Code CLI. Its `model`
argument is passed verbatim as `--model`: it does not use claudeR's API shortcut
resolver or API model validation. The CLI decides which aliases and model IDs are
available from its own version, account access, and configuration.

```{r}
claude_code("Review the files in this repository", model = "opus")
```

## Further Resources

- [Anthropic Models overview](https://platform.claude.com/docs/en/about-claude/models/overview)
- [Model IDs and versioning](https://platform.claude.com/docs/en/about-claude/models/model-ids-and-versions)
- [Anthropic API documentation](https://platform.claude.com/docs/en/api/overview)
- [claudeR GitHub repository](https://github.com/yrvelez/claudeR)
