---
title: "Getting Started with claudeR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with claudeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction
The claudeR package provides a convenient R interface to the Anthropic Claude API.
With claudeR, you can easily generate AI-powered text responses using Claude's
advanced language models without managing the underlying API request details.

## Prerequisites

Before using claudeR, you need:

1. An Anthropic API key (get one at <https://console.anthropic.com/>)
2. The API key set as an environment variable

### Setting your API key

The recommended way to set your API key is via an environment variable:

```{r}
Sys.setenv(ANTHROPIC_API_KEY = "your-api-key-here")
```

For persistent use, add this to your `.Renviron` file:

```
ANTHROPIC_API_KEY=your-api-key-here
```

## Basic Usage

Load the package:

```{r setup}
library(claudeR)
```

### Simple Query

For Claude 3 and newer models, prompts must be provided as a list of messages:
```{r}
response <- claudeR(
  prompt = list(list(role = "user", content = "What is the capital of France?")),
  model = "claude-sonnet-4-5-20250929",
  max_tokens = 100
)
print(response)
```

### Using a System Prompt

You can provide context or instructions using a system prompt:

```{r}
response <- claudeR(
  prompt = list(list(role = "user", content = "Translate: Hello, how are you?")),
  model = "claude-sonnet-4-5-20250929",
  system_prompt = "You are a Spanish translator. Respond only with the translation.",
  max_tokens = 100
)
print(response)
```

### Multi-turn Conversations

Build a conversation by including multiple messages:

```{r}
response <- claudeR(
  prompt = list(
    list(role = "user", content = "My name is Alice."),
    list(role = "assistant", content = "Hello Alice! Nice to meet you."),
    list(role = "user", content = "What's my name?")
  ),
  model = "claude-sonnet-4-5-20250929",
  max_tokens = 100
)
print(response)
```

## Extended Thinking Mode

Claude supports an extended thinking mode where it can "think through" complex
problems step by step before providing a response:

```{r}
response <- claudeR(
  prompt = list(list(role = "user", content = "Solve: What is 127 * 83?")),
  model = "claude-sonnet-4-5-20250929",
  max_tokens = 8000,
  thinking = list(type = "enabled", budget_tokens = 5000),
  return_thinking = TRUE
)

# View the thinking process
cat("Thinking:\n", response$thinking, "\n\n")

# View the final answer
cat("Response:\n", response$response)
```

## Available Models

claudeR supports multiple Claude model families:

- **Claude 4.5**: `claude-opus-4-5-20251101`, `claude-sonnet-4-5-20250929`, `claude-haiku-4-5-20251001`
- **Claude 3.5**: `claude-3-5-sonnet-20241022`, `claude-3-5-haiku-20241022`
- **Claude 3**: `claude-3-opus-20240229`, `claude-3-sonnet-20240229`, `claude-3-haiku-20240307`
- **Claude 2** (legacy): `claude-2.1`, `claude-2.0`

## Parameters

Key parameters for the `claudeR()` function:

| Parameter | Description | Default |
|-----------|-------------|---------|
| `prompt` | List of messages (Claude 3+) or string (Claude 2) | Required |
| `model` | Model identifier | `"claude-sonnet-4-5-20250929"` |
| `max_tokens` | Maximum tokens in response | `100` |
| `temperature` | Randomness (0-1) | `0.7` |
| `system_prompt` | System instructions | `NULL` |
| `thinking` | Enable extended thinking | `NULL` |
| `stream_thinking` | Stream output in real-time | `TRUE` |
| `return_thinking` | Include thinking in output | `FALSE` |

## Error Handling

If the API request fails, claudeR will return `NULL` and display a warning with
error details. Common issues include:

- Invalid or missing API key
- Exceeded rate limits
- Invalid model name
- Malformed prompt format

## Further Resources

- [Anthropic API Documentation](https://docs.anthropic.com/)
- [claudeR GitHub Repository](https://github.com/yrvelez/claudeR)
